---
title: "Initial Data Cleaning"
author: "Andomei Smit: SMTAND051"
date: "22/03/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
```

# Load packages

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(xtable)
library(splitstackshape)
library(FactoMineR)
library(ggplot2)
library(factoextra)
library(lattice)
library(permute)
library(vegan)
Sys.setenv(RGL_USE_NULL = TRUE) # need this to get RDRToolbox installed
library(RDRToolbox)
library(FNN)
library(e1071)
rm(list=ls())
```

# Data preparation

## Load data
```{r load_data}
# load original margins
margin_df <- read.delim("data/data_Mar_64.txt", sep = ",")
shape_df <- read.delim("data/data_Sha_64.txt", sep = ",")
texture_df <- read.delim("data/data_Tex_64.txt", sep = ",")

colnames(margin_df) <- c("Species", paste0("Feature_", 1:64))
colnames(shape_df) <- c("Species", paste0("Feature_", 1:64))
colnames(texture_df) <- c("Species", paste0("Feature_", 1:64))
```

Investigate the missing features and remove the species missing features. Note that the samples do not have a unique identifier that is the same for each sample across the three features. This means that for any sample missing one of the features, it is impossible to know which sample it relates to. For that reason, all the samples for that species needs to be removed.

## Remove species with samples missing features
```{r quick_clean}
# texture only has 1598 observations:
#table(texture_df$Species) # acer campestre only has 14 images
#table(shape_df$Species) # only has 15 Acer Capillipes

margin_df <- margin_df[-which(margin_df$Species== "Acer Campestre"),]
shape_df <- shape_df[-which(shape_df$Species== "Acer Campestre"),]
texture_df <- texture_df[-which(texture_df$Species== "Acer Campestre"),]

margin_df <- margin_df[-which(margin_df$Species== "Acer Capillipes"),]
shape_df <- shape_df[-which(shape_df$Species== "Acer Capillipes"),]
texture_df <- texture_df[-which(texture_df$Species== "Acer Capillipes"),]
```

## Summary statistics for feature vectors per feature
Generate some summary statistics to look at each of the feature vectors.

```{r summ_margin}
# Generate summary statistics for continuous variables and round to 2 decimals
summary_stats_margin <- margin_df %>%
  summarise(across(where(is.numeric), 
                   list(Count = ~sum(!is.na(.)),
                        Mean = ~round(mean(., na.rm = TRUE), 2),
                        SD = ~round(sd(., na.rm = TRUE), 2),
                        Min = ~round(min(., na.rm = TRUE), 2),
                        Q1 = ~round(quantile(., 0.25, na.rm = TRUE), 2),
                        Median = ~round(median(., na.rm = TRUE), 2),
                        Q3 = ~round(quantile(., 0.75, na.rm = TRUE), 2),
                        Max = ~round(max(., na.rm = TRUE), 2)),
                   .names = "{.col}_{.fn}"))

# Reshape data into a cleaner table
summary_stats_margin_table <- summary_stats_margin %>%
  pivot_longer(cols = everything(), names_to = "Variable_Statistic", values_to = "Value") %>%
  separate(Variable_Statistic, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%  # Correct splitting
  pivot_wider(names_from = Statistic, values_from = Value)

# Print the summary table
print(summary_stats_margin_table)

hist(summary_stats_margin_table$Mean)
hist(summary_stats_margin_table$SD)

range(summary_stats_margin_table$Mean)
```

```{r summ_texture}
# Generate summary statistics for continuous variables and round to 2 decimals
summary_stats_texture <- texture_df %>%
  summarise(across(where(is.numeric), 
                   list(Count = ~sum(!is.na(.)),
                        Mean = ~round(mean(., na.rm = TRUE), 2),
                        SD = ~round(sd(., na.rm = TRUE), 2),
                        Min = ~round(min(., na.rm = TRUE), 2),
                        Q1 = ~round(quantile(., 0.25, na.rm = TRUE), 2),
                        Median = ~round(median(., na.rm = TRUE), 2),
                        Q3 = ~round(quantile(., 0.75, na.rm = TRUE), 2),
                        Max = ~round(max(., na.rm = TRUE), 2)),
                   .names = "{.col}_{.fn}"))

# Reshape data into a cleaner table
summary_stats_texture_table <- summary_stats_texture %>%
  pivot_longer(cols = everything(), names_to = "Variable_Statistic", values_to = "Value") %>%
  separate(Variable_Statistic, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%  # Correct splitting
  pivot_wider(names_from = Statistic, values_from = Value)

# Print the summary table
print(summary_stats_texture_table)

hist(summary_stats_texture_table$Mean)
hist(summary_stats_texture_table$SD)

#range(summary_stats_margin_table$Mean)
```

```{r summ_shape}
# Generate summary statistics for continuous variables and round to 2 decimals
summary_stats_shape <- shape_df %>%
  summarise(across(where(is.numeric), 
                   list(Count = ~sum(!is.na(.)),
                        Mean = ~round(mean(., na.rm = TRUE), 2),
                        SD = ~round(sd(., na.rm = TRUE), 2),
                        Min = ~round(min(., na.rm = TRUE), 2),
                        Q1 = ~round(quantile(., 0.25, na.rm = TRUE), 2),
                        Median = ~round(median(., na.rm = TRUE), 2),
                        Q3 = ~round(quantile(., 0.75, na.rm = TRUE), 2),
                        Max = ~round(max(., na.rm = TRUE), 2)),
                   .names = "{.col}_{.fn}"))

# Reshape data into a cleaner table
summary_stats_shape_table <- summary_stats_shape %>%
  pivot_longer(cols = everything(), names_to = "Variable_Statistic", values_to = "Value") %>%
  separate(Variable_Statistic, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%  # Correct splitting
  pivot_wider(names_from = Statistic, values_from = Value)

# Print the summary table
print(summary_stats_shape_table)

hist(summary_stats_shape_table$Mean)
hist(summary_stats_shape_table$SD)

#range(summary_stats_margin_table$Mean)
```

Shape is not on the same measurement scale, thus all the feature vectors will be centered and scaled.

## Scale and center all data

```{r scale_center_date}
margin_df[,-1] <- scale(margin_df[,-1])
shape_df[,-1] <- scale(shape_df[,-1])
texture_df[,-1] <- scale(texture_df[,-1])

```

```{r echo=FALSE}
rm(summary_stats_margin)
rm(summary_stats_margin_table)
rm(summary_stats_shape)
rm(summary_stats_shape_table)
rm(summary_stats_texture)
rm(summary_stats_texture_table)
```

## Combine data from different features

```{r merge_data}
# make sure all species are in the same order:
## assign unique ID to sample:
## let i_j be sample j from sample i

check_order <- cbind(unique(texture_df$Species),
unique(shape_df$Species),
unique(margin_df$Species)) # these are not in the same order!

count_shape_species_switch <- 0
cur_species <- shape_df$Species[1]
for(i in 2:nrow(shape_df)){
  compare_species <- shape_df$Species[i]
  if(cur_species!= compare_species){
    count_shape_species_switch <- count_shape_species_switch +1
  }
  cur_species <- compare_species
}
# for shape the species only swap 97 times, meaning that the data lists all 16 samples of a species at a time

count_margin_species_switch <- 0
cur_species <- margin_df$Species[1]
for(i in 2:nrow(margin_df)){
  compare_species <- margin_df$Species[i]
  if(cur_species!= compare_species){
    count_margin_species_switch <- count_margin_species_switch +1
  }
  cur_species <- compare_species
}
# also only swaps 97 times

count_texture_species_switch <- 0
cur_species <- texture_df$Species[1]
for(i in 2:nrow(texture_df)){
  compare_species <- texture_df$Species[i]
  if(cur_species!= compare_species){
    count_texture_species_switch <- count_texture_species_switch +1
  }
  cur_species <- compare_species
}
# also swaps only 97 times

# number the samples:
margin_df$sample_num <- rep(1:16, 98)
shape_df$sample_num <- rep(1:16, 98)
texture_df$sample_num <- rep(1:16, 98)

# remove variables no longer needed for analysis
rm(count_margin_species_switch)
rm(count_shape_species_switch)
rm(count_texture_species_switch)
rm(compare_species)
rm(cur_species)
rm(i)
rm(check_order)

# merge on species and sample number:
## analysis ready dataset 1 (ard1)
ard1 <- merge(shape_df, margin_df, by = c("Species", "sample_num"))

ard1 <- merge(ard1, texture_df, by = c("Species", "sample_num"))

# add unique ID:
ard1$ID <- paste(ard1$Species, ard1$sample_num)
```

### Write data to csv

```{r clean_data_csv}
write.csv(ard1, "created_data/ard1.csv")
```

