---
title: "PCA and ISOMAP Dimension Reduction"
author: "Andomei Smit: SMTAND051"
date: "02/03/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
```

# Load packages

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(xtable)
library(splitstackshape)
library(FactoMineR)
library(ggplot2)
library(factoextra)
library(lattice)
library(permute)
library(vegan)
Sys.setenv(RGL_USE_NULL = TRUE) # need this to get RDRToolbox installed
library(RDRToolbox)
library(FNN)
library(e1071)
rm(list=ls())
```

# Load data

```{r load_data}
rm(list=ls())
ard1 <- read.csv("created_data/ard1.csv") # full cleaned data set (analysis ready data 1)

ard1_M <- read.csv("created_data/ard1_M.csv") # set for model building (train and validation)

ard1_T <- read.csv("created_data/ard1_T.csv") # set for model testing
```


# Dimension Reduction

## PCA

```{r}
# columns to ignore:
ignore_cols <- c("Species", "ID", "sample_num")
data.pca <- princomp(ard1_M[,-which(colnames(ard1_M)%in% ignore_cols)])
summary(data.pca)

fviz_eig(data.pca, addlabels = FALSE, barcolor = "darkgreen", barfill = "darkgreen", ncp = 15)


# cumulative variance: 
## 4pcs: 0.483197 - possible first elbow
## 5pcs: 0.51959
## 10pcs: 0.64974 - possible second elbow
## 20pcs: 0.77847
## 25pcs: 0.81676
## 30pcs: 0.84533 - large proportion of variance
```


```{r pca_csv}
pca.scores <- data.pca$scores # all scores
# write the transformed data from 4PCs to csv
write.csv(pca.scores[,1:4], "created_data/pca_4pcs_data.csv")

# write the transformed data from 10PCs to csv
write.csv(pca.scores[,1:10], "created_data/pca_10pcs_data.csv")

# write the transformed data from 30PCs to csv
write.csv(pca.scores[,1:30], "created_data/pca_30pcs_data.csv")
```

## ISOMAP
First find the smallest k such that the graph of nearest neighbours is fully connected.

### Find smallest viable value of k
```{r isomap_min_k}
# columns to ignore:
ignore_cols <- c("Species", "ID", "sample_num")
data_cols <- which(colnames(ard1_M)%in% ignore_cols)

# Find the smallest k that allows ISOMAP to run
k_values <- seq(5, 30, 1)  # Test k from 5 to 30
optimal_k <- NA

for (k in k_values) {
  tryCatch({
    isomap_result <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 2, k = k)
    optimal_k <- k  # If ISOMAP runs successfully, store k and break loop
    break
  }, error = function(e) {
    # If an error occurs, continue to next k
  })
}

# Print the first valid k
if (!is.na(optimal_k)) {
  print(paste("Optimal k (smallest fully connected graph):", optimal_k))
} else {
  print("None of the current values for k resulted in a fully connected graph. Increase k and retry.")
}
```

Smallest k is 13. Thus we can test a range of k from 13 to 30.

### Find optimal value of k and d using residual variance
In order to do this, we will fix the number of dimensions to use for MDS iteratively between 2 to 20 and iterate through possible values of K. For each value of k for a set dimension for the MDS, the residual variance will be calculated and stored. The aim is to find a value of k and corresponding dimensions for MDS that will minimize this variance.

In order to calculate the residual variance, we need the original geodesic distances, that is the sum of the euclidean distances along the shortest paths from the K-nn graphs.

```{r get_geodesic_dist}
# function to calculate the geodesic distances in the original high dimensional space

calc_geodestic_dist <- function(k, n = nrow(ard1_M)){
  # calculate the knn
  knearestn <- get.knn(ard1_M[,-data_cols], k = k)
  neighbor_index <- as.matrix(knearestn$nn.index) # index of neighbors
  neighbor_dist <- as.matrix(knearestn$nn.dist) # distance to neighbors
  
  # construct a matrix of all adjacency distances for i to j, if they are not reachable,
  # leave the matrix entry as NA:
  nn_distances <- matrix(NA, nrow = nrow(ard1_M), ncol = nrow(ard1_M))
  for(i in 1:n){
    for(j in 1:k){
      # for each observation i, find the index of it's neighbor j and 
      #save their distance there
      nn_distances[i,neighbor_index[i, j]] = neighbor_dist[i,j]
    }
  }
  
  # find the shortest path between all observations along these distances
  shortest_paths <- allShortestPaths(nn_distances)
  
  # initialise a matrix for geodesic distances
  geodesicdist <-  matrix(0,n,n)
  
  # find the exact path from each obs to all other observations and get the distance
  for (i in 1:n){
    for(j in 1:n){
      path <- extractPath(shortest_paths, i, j)
      total_dist = 0
      for(l in 1:(length(path)-1)){
       total_dist = total_dist +
         nn_distances[path[l],path[l+1]]
      }
      geodesicdist[i,j] = total_dist
    }
  }
  diag(geodesicdist) <- 0 # set diagonal to zero
  
  return(geodesicdist)
}
```

Next we need a function to calculate the residual variances.

```{r residual_var}
# function to calculate the residual variances

calc_residual_variance <- function(geodesic_distances, isomap_embedding) {
  # Compute Euclidean distances in ISOMAP-reduced space
  reduced_distances <- as.matrix(dist(isomap_embedding))

  # Flatten both matrices into vectors for correlation calculation
  geo_vector <- as.vector(geodesic_distances)
  euclid_vector <- as.vector(reduced_distances)

  # Compute R² (squared Pearson correlation)
  R2 <- cor(geo_vector, euclid_vector, use = "complete.obs")^2

  # Compute residual variance
  residual_variance <- 1 - R2

  return(residual_variance)
}

# test the function:
# get isomap for k=13 and d=2:
#isomap_k13_d2 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 2, k = 13)

#geodesic_k13 <- calc_geodestic_dist(13)

#calc_residual_variance(geodesic_k13, isomap_k13_d2$points)
```

Lastly, we need a function that puts all of this together.

```{r isomap_tuning}
# function that takes k and d (dimensions for MDS) and returns the residual variance
# between the distances of the original high dimensional data and the lower dimensional distances

isomap_tuning <- function(k_nn, d){
  # get the isomap configuration for the given k and d
  isomap_result <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = d, k = k_nn)
  
  # get the geodesic distances for the given k
  geodesic_dist <- calc_geodestic_dist(k_nn)
  
  # calculate the residual variance
  resid_var <- calc_residual_variance(geodesic_dist, isomap_result$points)
  
  return(resid_var)
}
```

```{r apply_isomap}
# get the results for fixed d
# Define values to test
k_values <- seq(13, 63, 5)   # Example values for k_nn
d_values <- seq(2, 102, 10)   # Example values for d

# Create all (k_nn, d) combinations
param_grid <- expand.grid(k_nn = k_values, d = d_values)

# Apply isomap_tuning() to all (k_nn, d) pairs using mapply
residual_variances <- mapply(isomap_tuning, param_grid$k_nn, param_grid$d)

# Store results in a data frame
results_df <- cbind(param_grid, residual_variance = residual_variances)

# View results
print(results_df)

# save these results
write.csv(results_df,"created_data/isomap_tuning_results.csv")
```


```{r isomap_plots}
# Plot residual variance vs. k (lines for d)
plot_k_vs_residual <- ggplot(results_df, aes(x = k_nn, y = residual_variance, color = as.factor(d))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Residual Variance vs. k (for Different d Values)",
       x = "Number of Neighbors (k)",
       y = "Residual Variance",
       color = "Dimensions (d)") +
  theme_minimal() +
  theme(legend.position = "right")

# Plot residual variance vs. d (lines for k)
plot_d_vs_residual <- ggplot(results_df, aes(x = d, y = residual_variance, color = as.factor(k_nn))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Residual Variance vs. d (for Different k Values)",
       x = "Number of Dimensions (d)",
       y = "Residual Variance",
       color = "Neighbors (k)") +
  theme_minimal() +
  theme(legend.position = "right")

# Show both plots
print(plot_k_vs_residual)
print(plot_d_vs_residual)

# Save the first plot (Residual Variance vs. k, different d values)
ggsave("plots/residual_vs_k.pdf", plot = plot_k_vs_residual, width = 6, height = 4)

# Save the second plot (Residual Variance vs. d, different k values)
ggsave("plots/residual_vs_d.pdf", plot = plot_d_vs_residual, width = 6, height = 4)
```

From the plot that separates the lines based on dimensions, we see that the number of dimensions does not differentiate a reduction in residual variance past 12. I.e. from dimensions 12 onwards, there is not much gain in reduction of residual variance for higher dimensions as compared to d = 12. Setting the number of dimensions to 12 and trying to find a suitable value for k, it can be seen that there is only a slight reduction in variance for k equal to 13, 18 and 23. These are 0.48, 0.43 and 0.4, respectively. However, the final decision will be left after the clustering algorithm has been applied.

```{r isomap_resid_var_calc}
# residual variance for k = 13 and d = 12
results_df$residual_variance[which(results_df$k_nn== 13 & results_df$d==12)] # 0.4794

# residual variance for k = 18 and d = 12
results_df$residual_variance[which(results_df$k_nn== 18 & results_df$d==12)] # 0.4344

# residual variance for k = 23 and d = 12
results_df$residual_variance[which(results_df$k_nn== 23 & results_df$d==12)] # 0.4025
```

```{r isomap_csv}
# isomap with k = 13 and d = 12
isomap_k13_d12 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 12, k = 13)
write.csv(isomap_k13_d12$points, "created_data/isomap_k13_d12.csv")

# isomap with k = 18 and d = 12
isomap_k18_d12 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 12, k = 18)
write.csv(isomap_k18_d12$points, "created_data/isomap_k18_d12.csv")

# isomap with k = 23 and d = 12
isomap_k23_d12 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 12, k = 23)
write.csv(isomap_k23_d12$points, "created_data/isomap_k23_d12.csv")

rm(isomap_k13_d12)
rm(isomap_k18_d12)
rm(isomap_k23_d12)
```

Trying with d= 22 as well.

```{r isomap_csv}
# isomap with k = 13 and d = 22
isomap_k13_d22 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 22, k = 13)
write.csv(isomap_k13_d22$points, "created_data/isomap_k13_d22.csv")

# isomap with k = 18 and d = 12
isomap_k18_d22 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 22, k = 18)
write.csv(isomap_k18_d22$points, "created_data/isomap_k18_d22.csv")

# isomap with k = 23 and d = 12
isomap_k23_d22 <- isomap(dist(as.matrix(ard1_M[,-data_cols])), ndim = 22, k = 23)
write.csv(isomap_k23_d22$points, "created_data/isomap_k23_d22.csv")

rm(isomap_k13_d22)
rm(isomap_k18_d22)
rm(isomap_k23_d22)
```

```{r isomap_clear_workspace}
rm(geodesic_k13)
rm(param_grid)
rm(isomap_k13_d2)
rm(plot_d_vs_residual)
rm(plot_k_vs_residual)
rm(results_df)
rm(d)
rm(d_values)
rm(k)
rm(k_values)
rm(n)
rm(optimal_k)
rm(residual_variances)
rm(calc_geodestic_dist)
rm(calc_residual_variance)
rm(isomap_tuning)
```
